### 2020-04-15 dataclean.tex
updated dataclean

% !TEX root = main.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Viewing and Cleaning the Data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Viewing and Cleaning the Data}
\label{section:dataclean}
As we need to recover durations from this data, correct processing
involves time adjustments for switching between daylight saving time
and standard time. We perform this by setting a reference time zone
(Eastern time) and converting all date-times from strings into
date-time variables \fix{(is there a name for this?)} with the R
\pkg{lubridate} package \cite{lubridate}, which enables appropriate
date arithmetic.

\fix{Other processing:} enable missing values, and create some new
variables. The event is “life” when both insert and remove are
present, “life0” when insert = remove, otherwise it is whatever string
is in insert, which is “DBE” or “Off The BUS”. Also separate location
into its components and fill in (repeat) serial numbers and locations
for all records.

Reduce to only “life” records and mark them with ending DBE, OTB,
“out”, or none events. The result is just insert and remove with event
marks to indicate DBE or OTB or “out” (last seen). bad indicates that
a DBE or OTB occurred yet the GPU was not taken out. For a first cut,
keep the bad ones in. Also keep duration to later aggregate into full
life times.

Next, aggregate into one record per serial number with total life time
and first insert time. For proper censoring treatment: Indicate if
event occurred or still in service: out = fail, dbe = fail, otb =
fail, NA = right censored (still in service). Add some other
variables, such as GPU in single location or moved one or more times,
mark new\_batch group, etc. to use in modeling later.

To get some intuition for the GPU lifetimes on Titan, we give two views of
a hundred GPUs (Figure~\ref{fig:gpuview}) and a hundred locations
(Figure~\ref{fig:locview}). The GPU view is ``by serial number'' that
shows when each GPU was installed and removed at various locations as
well as its DBE and OTB events. The second view by location, shows
when different GPUs were installed and removed at a location, and
their OTB and DBE events.
\begin{figure*}[tb]
  \includegraphics[width=6.5in]{figs/sample_sn.pdf}
  \caption{GPU serial number view of life and failures. Legend: black
    dots are installs, black lines are lifetimes at installed
    location, blue squares are OTB events, red triangles are DBE
    events, and black ] are ``last seen'' events.}
  \label{fig:gpuview}
\end{figure*}
\begin{figure*}[tb]
  \includegraphics[width=6.5in]{figs/sample_loc.pdf}
  \caption{GPU location view of life and failures.  Legend: black
    dots are installs, black lines are lifetimes at installed
    location, blue squares are OTB events, red triangles are DBE
    events, and black ] are ``last seen'' events.}
  \label{fig:locview}
\end{figure*}
The serial number view clearly shows the two batches and the more
frequent OTB and DBE events in the old batch. It is also clear from
this view that most new GPUs stay at the initial install location
whereas the old GPUs were occasionally reinstalled at new locations.
The location view shows that each location was operational almost all
the time with small gaps when GPUs were changed out.

\fix{Were the GPUs swapped out as a blade or were individual GPUs
  swapped on a blade?}

>>>>

% !TEX root = main.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Viewing and Cleaning the Data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Viewing and Cleaning the Data}
\label{section:dataclean}
As we need to recover durations in operation from this data, correct
processing involves time adjustments for switching between daylight
saving time and standard time and leap years. We perform this by
setting a reference time zone (Eastern time) and converting all
date-times from strings into POSIX date-time variables with the R
\pkg{lubridate} package \cite{lubridate}, which enables appropriate
date arithmetic and sensible date constructs for graphs.

As most analysis software uses table-like data, we fill the needed
repeats of values missing in the raw data (see
Fig.~\ref{fig:dataraw}. First, we reduce the data to the GPU that were
removed later than 2013 to focus our analysis epoch on units installed
in the second rework cycle. After this reduction, there were stil six
older units remaining, which we also removed to have a clean third
epoch set of units for the analysis. We do some further processing to
handle time overlaps in a tiny fraction (under 0.0007 in GPU life and
0.0002 in location life) of recorded life in the raw records by simply
dropping the overlapping life times.

Next, we aggregate into one record per serial number with a total life
time the first insert time, the last remove time, and a number of
other quantities such as location where the longest time is spent, the
proportion of time at the longest location, and the number of DBE or
OTB events.

To get some intuition for the GPU lifetimes on Titan, we give two
views of 80 randomly selected GPUs (Figure~\ref{fig:gpuview}) and 80
randomly selected locations (Figure~\ref{fig:locview}). The GPU view
visually documents the life of the unit: when it was installed and
removed at various locations, its DBE and OTB events, and the last
time it was seen. The location view documents the life of a location:
when different GPUs were installed and removed, their OTB and DBE
events, and whether a removal was the last time the unit was seen on
the system. These views were critical to understanding the data and to
verifying various processing decisions.
\begin{figure*}[tb]
  \includegraphics[width=6.5in]{figs/sample_sn.pdf}
  \caption{GPU serial number view of life and failures. Legend: black
    dots are installs, black lines are lifetimes at installed
    location, blue squares are OTB events, red triangles are DBE
    events, and black ] are ``last seen'' events.}
  \label{fig:gpuview}
\end{figure*}
\begin{figure*}[tb]
  \includegraphics[width=6.5in]{figs/sample_loc.pdf}
  \caption{GPU location view of life and failures.  Legend: black
    dots are installs, black lines are lifetimes at installed
    location, blue squares are OTB events, red triangles are DBE
    events, and black ] are ``last seen'' events.}
  \label{fig:locview}
\end{figure*}

The third rework cycle was used to label GPUs as old batch and new
batch. The two batches are clearly identifiable in the GPU view. Also,
more frequent OTB and DBE events are apparent in the old batch. It is
also clear from this view that most new GPUs stay at the initial
install location whereas the old GPUs were occasionally reinstalled at
new locations. Nevertheless, a separate analysis determined that the
vast majority of time of the vast majority of units is spent at one
location by both the new and the old units.

The location view shows that each location was operational almost all
the time with small gaps when GPUs were changed out.

\fix{Were the GPUs swapped out as a blade or were individual GPUs
  swapped on a blade? Seems it was the full blade - this may further
  refine censoring. A DBE on one GPU triggers the replacement of the
  blade with the other three good gpus?}
